### 2024-02-27 Chapter5.tex
Added tissue optics section to intro

\section{Introduction}
As described in Chapter \ref{chap:intro}, hyperspectral imaging (HSI) is able to capture spatially resolved, diffuse reflectance spectra via non-contact imaging~\citep{Lu2014,Giannoni2018,Calin2014,Shapey2019}, which could enable quantitative, spatially-resolved, tissue oxygenation ($StO_2$) extraction. Many methods have been considered for this ranging from two or three wavelength approaches~\citep{Clancy2020} to wavelength-dependent analytical models including those investigated in Chapters \ref{chap:1layer} and \ref{chap:2layer} as these have been some of the prominent methods of doing so~\citep{Yudovsky2009, Jacques1999, Clancy2015, Clancy2020}. Other methods include deep-learning based approaches, however these rely on models or simulation to generate their training data~\citep{Wirkert2016}, or models based on the derivatives of the spectra, however these rely on extensive calibration and assume that the scattering contribution is wavelength-independent which may not be true for all tissues~\citep{Holmer2018}.

>>>>

\section{Introduction}
As described in Chapter \ref{chap:intro}, hyperspectral imaging (HSI) is able to capture spatially resolved, diffuse reflectance spectra via non-contact imaging~\citep{Lu2014,Giannoni2018,Calin2014,Shapey2019}, which could enable quantitative, spatially-resolved, tissue oxygenation ($StO_2$) extraction. Many methods have been considered for this ranging from two or three wavelength approaches~\citep{MacKenzie2018} to wavelength-dependent analytical models including those investigated in Chapters \ref{chap:1layer} and \ref{chap:2layer} as these have been some of the prominent methods of doing so~\citep{Yudovsky2009, Jacques1999, Clancy2015, Clancy2020}. Other methods include deep-learning based approaches, however these rely on models or simulation to generate their training data~\citep{Wirkert2016}, or models based on the derivatives of the spectra, however these rely on extensive calibration and assume that the scattering contribution is wavelength-independent which may not be true for all tissues~\citep{Holmer2018}.

### 2024-02-27 Introduction.tex
Added tissue optics section to intro

This chapter presents some clinical context of brain tumours (Section \ref{sec:braintumours}) and aneurysms (Section \ref{sec:aneurysms}) for which this research is targeted alongside the current, relevant, available clinical adjuncts. This is followed by a description of oxygen saturation (Section \ref{sec:oxygensat}) and some key existing technologies to compute this and their limitations (Section \ref{sec:stateofart}). Finally, the structure of this thesis is presented in Section \ref{sec:thesisoutline}.

@@

Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from tissue diffuse reflectance spectra. These range from the simplest ratiometric two-wavelength methods to more sophisticated analytical models~\citep{MacKenzie2018} or deep-learning based approaches~\citep{Ayala2023}. Using two or three wavelength models can provide good results but require data to be captured at these precise wavelengths, whereas analytical models can be applied to ranges of wavelengths for which they are developed. These analytical models also enable more detailed modelling of tissue including consideration of melanin when imaging through skin. Alternatively, Monte Carlo methods are an established approach to modelling light-tissue interaction for a range of wavelengths. It is, however, computationally challenging to apply in an inverse problem setting making the estimation of optical properties from tissue spectra non trivial on the basis of Monte Carlo simulations alone. To obtain the diffuse reflectance spectra intra-operatively, these techniques primarily use HSI which enable non-contact imaging with minimal change to surgical workflow.

\section{Current state of the art of $StO_2$ measurement}\label{sec:stateofart}
The challenges presented in current neurosurgical practice present an interesting area for development. Many technologies are being developed to address the need for $StO_2$ assessment intra-operatively for these applications. In this section some key research technologies are presented alongside their limitations and the choice to pursue HSI in this work is justified.

@@

\subsection{Probe-based diffuse reflectance spectroscopy}
When light is incident on a surface it can either be directly reflected in a specular (mirror-like) manner, or it can be internally scattered or absorbed before leaving the sample diffusely, as seen in Figure \ref{fig:diffuseR}. Specularly scattered light can carry topological information of the sample, this is examined using laser speckle spectroscopy which can investigate blood flow~\citep{Dunn2012}, however diffusely scattered light is linked to the absorption and scattering properties of the sample. 
\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{Tissue_figure.pdf}
    \caption{Schematic of specularly and diffusely reflected light.}
    \label{fig:diffuseR}
\end{figure}
Diffuse reflectance spectroscopy measures this diffusely scattered light from a sample in order to gain information on the chemical composition of the sample by investigating the absorbance and scattering properties of the sample. Whilst this work focuses on HSI as a method of diffuse reflectance spectroscopy, there are many fibre-optic probe-based systems for these measurements. These systems provide highly accurate diffuse reflectance spectra in a point-based measurement. Probe-based methods are able to provide high spectral resolution measurements with short acquisition times, whereas HSI usually requires longer acquisition periods for similar spectral quality~\citep{Dinish2017}. This high spectral resolution of probe-based systems comes at the cost of spatial resolution. Whilst measurements from multiple probe locations can be stitched together~\citep{Thrapp2020}, wide-field imaging is not possible~\citep{Nishidate2015}. This technique traditionally involves contact between the probe and the sample, however this contact can alter optical properties of the tissue sample~\citep{Miller2017}. To mitigate this, flexible probes have been proposed~\citep{Miller2017} or non-contact solutions have been developed which make use of lenses and remove the impact of specular reflections using polarising filters~\citep{Bish2011, Zhu2012}. %None of these measurement configurations, however, provide wide-field information. 
Diffuse reflectance spectroscopy has shown potential for tissue differentiation~\citep{Skyrman2022} and $StO_2$ measurement~\citep{Fredriksson2020} using methods that are also applied to HSI and so are discussed more thoroughly in the following section \ref{sec:introHSI}. These techniques are able to provide real-time, intra-operative, highly accurate measurements which show promise for tissue differentiation and $StO_2$ measurement, however they do not provide wide-field imaging and non-contact configurations introduce significantly more bulk to the system. For this reason HSI is investigated to overcome these limitations.

>>>>

This chapter presents some clinical context of brain tumours (Section \ref{sec:braintumours}) and aneurysms (Section \ref{sec:aneurysms}) for which this research is targeted alongside the current, relevant, available clinical adjuncts. This is followed by a description of oxygen saturation (Section \ref{sec:oxygensat}) and optical properties of tissues (Section \ref{sec:introoptics} alongside some key methods to calculate oxygen saturation using optical measurements. This is followed by some key existing technologies used for measurement of tissue and their limitations (Section \ref{sec:stateofart}). Finally, the structure of this thesis is presented in Section \ref{sec:thesisoutline}.

@@

%Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from tissue diffuse reflectance spectra. These range from the simplest ratiometric two-wavelength methods to more sophisticated analytical models~\citep{MacKenzie2018} or deep-learning based approaches~\citep{Ayala2023}. Using two or three wavelength models can provide good results but require data to be captured at these precise wavelengths, whereas analytical models can be applied to ranges of wavelengths for which they are developed. These analytical models also enable more detailed modelling of tissue including consideration of melanin when imaging through skin. Alternatively, Monte Carlo methods are an established approach to modelling light-tissue interaction for a range of wavelengths. It is, however, computationally challenging to apply in an inverse problem setting making the estimation of optical properties from tissue spectra non trivial on the basis of Monte Carlo simulations alone. To obtain the diffuse reflectance spectra intra-operatively, these techniques primarily use HSI which enable non-contact imaging with minimal change to surgical workflow. 
Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from optical measurements of tissues, including diffuse reflectance spectra. These models relate the extinction coefficients of haemoglobin to the optical properties of tissues to enable $StO_2$ calculation.

\section{Optical properties of tissues}\label{sec:introoptics}
To extract useful information, such as $StO_2$, from optical measurements of tissues it is necessary to be able to model the optical properties of these tissues. The magnitude of the absorption and scattering of light through tissue is characterised by an absorption coefficient $\mu_a$ and scattering coefficient $\mu_s$. Scattering can occur isotropically, the light wave has an equal probability of propagating in any direction from the point of scatter, or anisotropically, where the light wave is more likely to propogate in a direction either forwards or backwards from the point of scatter~\citep{Jacques2013}. This directional component of scatter is described by the anisotropy factor $g$ for bulk tissue, where 0 represents isotropic scatter and 1 or -1 represents perfect forwards or backwards scattering respectively~\citep{Jacques2013}. These scattering parameters are often combined into a single scattering measure $\mu_s'$ termed the reduced scattering coefficient according to Equation \eqref{eq:reducedRintro}~\citep{Jacques2013}. 
\begin{equation}
    \mu_s' = \mu_s(1-g)
    \label{eq:reducedRintro}
\end{equation}

$\mu_a$ is well defined by combining the concentrations ($c$) and extinction coefficients ($\epsilon$) of each chromophore ($i$) or combining the volume fraction ($f$) and the absorption coefficient ($\mu_a$) of each pure chromophore ($i$) in the sample at any wavelength ($\lambda$)~\citep{Jacques2013, MacKenzie2018}. 
\begin{equation}
    \mu_a(\lambda) = \sum_i c_i \epsilon_i(\lambda) = \sum_i f_i \mu_{a, i}(\lambda)
    \label{eq:definemua}
\end{equation}
For non-scattering media this $\mu_a$ can easily be converted to an absorbance measurement ($A$) by multiplying with the path length ($l$) of the sample through which the measurement is obtained according to the Beer-Lambert law~\cite{MacKenzie2018}. This also allows simple calculations to solve the inverse problem of calculating concentrations of chromophores from absorbance measurements of samples. 
\begin{equation}
    A = \mu_al
\end{equation}

Scattering behaviour of particles in biological tissues can be broadly divided into Rayleigh, Mie, and Raman scattering. Commonly, Rayleigh and Mie scattering are defined as the elastic scattering behaviour by particles significantly smaller or particles comparable to or larger than the incident wavelength of light respectively. More precisely, Mie scattering refers to the scattering of spheres of any size with Rayleigh scattering being the Rayleigh limit of Mie scattering for particles significantly smaller than the wavelength of incident light~\cite{Jacques2013}. Raman scattering refers to the inelastic scattering of light, however this effect is typically much smaller and is not easy to detect without long integration times for typical optical measurement systems~\citep{Kong2015}. Within the common measurement range of 400-1000nm, modelling Mie scattering alone is sufficient, however beyond this range a combination of Rayleigh and Mie scattering may be required to provide a good model of the scattering behaviour~\cite{Jacques2013}. This Rayleigh component at extreme wavelengths can be more significant for tissues with higher levels of collagen, whereas Mie theory continues to provide a good model of scattering for tissues where the predominant scatterer is fat~\cite{Jacques2013}. By using Mie theory, each of the scattering coefficients can be calculated using the diameter and volume fraction of the scattering particles for each wavelength of light~\citep{Hergert2012}. This information, however, is usually unknown for biological tissue and so the Mie scattering is often simplified to Equation \eqref{eq:Mieintro} where the reduced scattering at 500nm is taken as a reference point~\citep{Jacques2013}. 
\begin{equation}
    \mu_s' = a(\frac{\lambda}{500nm})^{-b}
    \label{eq:Mieintro}
\end{equation}

\subsection{Major $StO_2$ approaches}
By considering these optical properties of tissues, a variety of methods to extract $StO_2$ from measurements of tissue have been developed. In this section we will discuss some of the key methods used to extract $StO_2$ when using the range of wavelengths 400-1000nm to obtain in vivo measurements.

The most simplistic method of $StO_2$ estimation was the two-wavelength approach based on a pure Beer-Lambert model of tissue. This method assumes only oxygenated and deoxygenated haemoglobin ($HbO_2$ and $Hb$) as chromophores in tissue. 
\begin{equation}
    \mu_a = c_{Hb}\epsilon_{Hb} + c_{HbO_2}\epsilon_{HbO_2}
\end{equation}
In this method measurements of tissue are acquired at two wavelengths, one of which is an isosbestic point for the haemoglobin chromophore where the oxygenated and deoxygenated extinction coefficients are equal~\citep{MacKenzie2018}. This approach required significant calibration to account for scattering in tissues~\citep{MacKenzie2018}. Pulse oximetry is based on this approach, however the pulsatile nature of arterial blood allows the background non-pulsatile absorbance to be removed meaning that arterial $SpO_2$ can be calculated for the pulsatile portion of the signal only~\citep{MacKenzie2018}. This, however, is still confounded by the presence of other significant absorbers in high concentrations such as melanin or by low blood flow due to low temperature~\citep{Bangash2022, MacKenzie2018}. This allowed development of the three-wavelength approach. In this case, two isosbestic wavelengths are measured to calculate a wavelength-independent scattering coefficient with a third used for $StO_2$ extraction~\cite{MacKenzie2018}. Whilst this provides advantages over the two-wavelength approach, it has largely been superseded by hyperspectral approaches due to the limited number of appropriate wavelengths that can be used which satisfy the approximation of constant scattering and by the continued requirement for calibration with samples of known $StO_2$~\cite{MacKenzie2018}. By using ranges of wavelengths and investigating the derivative of the spectrum a hyperspectral alternative has been proposed, however this also assumes wavelength-independent scattering and requires significant calibration~\citep{Holmer2018}. To advance this field, Monte Carlo simulations were used to probabilistically predict the spectra that could be obtained from a sample with known optical properties~\cite{Wang1995}. These have been used to improve approaches based on the Beer-Lambert models by including differential path length factors which are fitted to simulated data~\cite{Clancy2020}.

Whilst Monte Carlo simulations provide reliable predictions of diffuse reflectance spectra, the inverse solution is computationally challenging making the estimation of optical properties from tissue spectra non trivial on the basis of Monte Carlo simulations alone. To make use of this, Monte Carlo based look-up tables have been used in some cases~\cite{Nguyen2021, Yang2022}. Alternatively, Monte Carlo is often used to validate or inform complex analytical models such as those investigated in this work by \citet{Jacques1999} and \citet{Yudovsky2009} which are discussed in detail in later chapters. Finally, deep learning algorithms have been trained using Monte Carlo simulated data and applied in vivo~\cite{Wirkert2017, Wirkert2016}. Whilst these algorithms hold promise for improving the accuracy of $StO_2$ extraction, they rely on the accurate modelling of the input $\mu_a$ and $\mu_s$ parameters for the Monte Carlo simulations and often do not show validation with measured data of known ground-truth. As an initial step to understanding and improving these approaches, the analytical models are investigated further in this work.

\section{Current state of the art of $StO_2$ measurement}\label{sec:stateofart}
The challenges presented in current neurosurgical practice present an interesting area for development with many possible methods of calculating $StO_2$ from in vivo measurements. Many technologies are being developed to address the need for $StO_2$ assessment intra-operatively for these applications. In this section some key research technologies are presented alongside their limitations and the choice to pursue HSI in this work is justified.

@@

\subsection{Probe-based diffuse reflectance spectroscopy}
When light is incident on a surface it can either be directly reflected in a specular (mirror-like) manner, or it can be internally scattered or absorbed before leaving the sample diffusely, as seen in Figure \ref{fig:diffuseR}. Specularly scattered light can carry topological information of the sample, this is examined using laser speckle spectroscopy which can investigate blood flow~\citep{Dunn2012}, however diffusely scattered light is linked to the absorption and scattering properties of the sample. 
\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{Tissue_figure.pdf}
    \caption{Schematic of specularly and diffusely reflected light.}
    \label{fig:diffuseR}
\end{figure}
Diffuse reflectance spectroscopy measures this diffusely scattered light from a sample in order to gain information on the chemical composition of the sample by investigating the absorbance and scattering properties of the sample. Whilst this work focuses on HSI as a method of diffuse reflectance spectroscopy, there are many fibre-optic probe-based systems for these measurements. These systems provide highly accurate diffuse reflectance spectra in a point-based measurement. Probe-based methods are able to provide high spectral resolution measurements with short acquisition times, whereas HSI usually requires longer acquisition periods for similar spectral quality~\citep{Dinish2017}. This high spectral resolution of probe-based systems comes at the cost of spatial resolution. Whilst measurements from multiple probe locations can be stitched together~\citep{Thrapp2020}, wide-field imaging is not possible~\citep{Nishidate2015}. This technique traditionally involves contact between the probe and the sample, however this contact can alter optical properties of the tissue sample~\citep{Miller2017}. To mitigate this, flexible probes have been proposed~\citep{Miller2017} or non-contact solutions have been developed which make use of lenses and remove the impact of specular reflections using polarising filters~\citep{Bish2011, Zhu2012}. %None of these measurement configurations, however, provide wide-field information. 
Diffuse reflectance spectroscopy has shown potential for tissue differentiation~\citep{Skyrman2022} and $StO_2$ measurement~\citep{Fredriksson2020} using methods that are also applied to HSI and are discussed more thoroughly in Sections \ref{sec:introoptics} and \ref{sec:introHSI}. These techniques are able to provide real-time, intra-operative, highly accurate measurements which show promise for tissue differentiation and $StO_2$ measurement, however they do not provide wide-field imaging and non-contact configurations introduce significantly more bulk to the system. For this reason HSI is investigated to overcome these limitations.
